---
title: 'Practical Machine Learning: Course Project'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---

## Synopsis

In this project, we analyse data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. Our goal is to apply machine learning algorithm to predict the manner in which they do exercise, i.e. how well (in the 5 classes) they will do it.

More information is available from the website here:  http://groupware.les.inf.puc-rio.br/har

## Data Processing

### Download and import data

First, we check whether the data files are already in the working directory and download the data files if necessary.

```{r}

# set options 
options(warn=-1);

# install.package("R.utils")
library(R.utils, quietly=TRUE);

# Download data
if (!file.exists("pml-training.csv")){
      download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", "pml-training.csv")
      }

if (!file.exists("pml-testing.csv")){
      download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", "pml-testing.csv")
      }

# Import data 
pml_training <- read.csv(file = 'pml-training.csv',
                         na.strings = c('NA',''))
pml_testing <- read.csv(file = 'pml-testing.csv', 
                        na.strings = c('NA',''))

```

### Preprocess the data

Take an initial look of the data:
```{r}
dim(pml_training)
dim(pml_testing)
```

The training data contains 19622 records of 160 features. The testing data contains 20 sets of data for further evaluation of our developed model.

Test the number of NAs in each features and see whether the feature have sufficient records to be included in the analysis:
```{r}
na_test1 = sapply(pml_training, function(x) {sum(is.na(x))})
table(na_test1)

na_test2 = sapply(pml_testing, function(x) {sum(is.na(x))})
table(na_test2)
```

There are 100 features with lots of NAs (> 19000 out of the 19622 records in the training data; and 20 out of the 20 records in the testing data). Remove these features:
```{r}
# Get the features with lots of NAs
NAs1 = names(na_test1[na_test1>19000])
NAs2 = names(na_test2[na_test2==20])
# Make sure the features are the same 
NAs1 == NAs2
# Remove those features with lots of NAs
pml_training = pml_training[, !names(pml_training) %in% NAs1]
pml_testing = pml_testing[, !names(pml_testing) %in% NAs2]
```

Take a look of the resulting features:
```{r}
names(pml_training)
```

The first seven features are not of our interest - not related to movement, thus can be removed.
```{r}
pml_training = pml_training[,-c(1:7)]
pml_testing = pml_testing[,-c(1:7)]
dim(pml_training)
dim(pml_testing)
```

## Model Development

### Select a model
The variable we will be predicting on is classe:
```{r}
summary(pml_training$classe)
library(ggplot2, quietly = TRUE)
ggplot(pml_training) + geom_histogram(aes(x = classe, fill= classe))

```

The classes are quit well balanced (of the same order of 10^3), a random forest model can be used.

### Build a model
Load the necessay packages:
```{r}
library(caret, quietly=TRUE);library(e1071, quietly=TRUE);
library(randomForest, quietly=TRUE);

```

Partition the training data to training set and testing set:
```{r}
set.seed(123)
inTrain <- createDataPartition(pml_training$classe, p=0.75, list=FALSE )
training <- pml_training[inTrain,]
testing <- pml_training[-inTrain,]
dim(training)
dim(testing)
```

Fit a random forest model, use doParallel package to run on 4 cores.
```{r}
# Load packages for parallel computing
library(doParallel, quietly=TRUE);
library(Hmisc, quietly=TRUE);

set.seed(123)
# run on 4 cores, 200 trees each
registerDoParallel()
mod_rf <- foreach(ntree=rep(200,4), .combine=randomForest::combine,
                  .packages="randomForest") %dopar% {
                        randomForest(classe ~ ., data = training,
                                     ntree=ntree)
                        }
```

### Model Evaluation
Use the testing set from data partition to evaluate the model developed.

```{r}
predictions1 <- predict(mod_rf, training)
confusionMatrix(predictions1,training$classe)

predictions2 <- predict(mod_rf, testing)
confusionMatrix(predictions2,testing$classe)
```

The prediction accuracy on the test data set (99.5%) is greater than 99%. The out of sample error is less than 1% (1- testing accuracy = 0.5%). The model developed performs very well.

## Further Prediction and Submission
Get the prediction for the testing cases:

```{r}
ptest <- predict(mod_rf, pml_testing)

ptest

```

Write text files for submission:
```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(ptest)

```

## Conclusions
A model is developed to predict how well the participants will do exercise based on their movement data. The out of sample error is estimated to be 0.5%, which is very promising. 


Last Modified: 27 Sep 2015

